Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-31 paper2.qmd
section:theoretical; process:write,scms,dags

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The study then addresses these issues by extending Thurstone's model through a systematic and integrated approach based on causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's model to address these challenges. The model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. A thought experiment can illustrate this concept. In this experiment, the researcher observes the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. In this scenario, @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ. Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area)

@@

<!-- 4. segway for next section -->
In aggregate, the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates. This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp. 2]. Consequently, addressing these issues by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.

@@

<!-- recording: Sven and Tine 25.01.22; time: 00:05:40 - 00:06:30 -->
<!-- recording: Sven and Tine 25.01.22; time: 00:08:25 - 00:31:56 -->

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
For instance, researchers might aim to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method. Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text. In this scenario, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Researchers would then use estimates from the BTL model, or its transformations, to approximate this estimand.

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents strong evidence that the BTL model has several statistical and measurement limitations. These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one presented in the previous paragraph. Identification is crucial because it is a necessary condition for ensuring consistent estimators. Without identification, researchers cannot achieve consistency even with "infinite" and error-free data. In other words, the estimates derived from the estimator fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp. 5].

@@

<!-- 5. What the section will do? -->
As a result, this section addresses the identified limitations by extending Thurstone's model through a systematic and integrated approach based on the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments. These features include the selection of judges, stimuli, and comparisons. Combining these components clarifies the interactions among all actors and processes involved in CJ experiments. Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into a *conceptual-population model*. Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into a *sample-comparison model*.

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, it is important to define SCMs and DAGs. SCMs are formal mathematical models defined by a set of *endogenous* variables, a set of *exogenous* variables, and a set of functions [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

For instance, following the scenario outline in @sec-theoretical, @fig-cj04_scm displays the SCM that relates the outcome of the CJ experiment $(O_{abhijk})$, to several endogenous variables $V$. These variables include the discriminal difference $(D_{abhijk})$, perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$, formally expressed as $V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$. The structural equations $\{f_{O},f_{D},f_{PT}\}$ define the dependencies among these variables. Notably, in this scenario, the judges produce a dichotomous outcome $(O_{abhijk}) = \{0,1\}$. Moreover, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ represents the comparison number. Regarding the $k$ index, the index accounts for cases in which judges perform multiple comparisons of the same texts written by the same students, as the one discussed in @sec-theoretical.

<!-- 2. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables. The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- The (latent) discriminal difference of the stimuli directly determines the (manifest) outcome of the pairwise comparisons -->

<!-- The (latent) "perceived" discriminal processes for the stimuli directly determines their discriminal difference -->

<!-- The (latent) "true" discriminal processes for the stimuli and the judges' biases directly determines their (latent) "perceived" discriminal processes -->

@@

Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$. In this scenario, the figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The study then addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The new model integrates the theoretical core principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. A thought experiment can illustrate this concept. In this experiment, the researcher observes the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ. Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area)

@@

<!-- 4. segway for next section -->
In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates. This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp. 2]. Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits. For this reason, the study employs the structural approach to causal inference, as it fulfills these requirements.

@@

<!-- recording: Sven and Tine 25.01.22; time: 00:05:40 - 00:06:30 -->
<!-- recording: Sven and Tine 25.01.22; time: 00:08:25 - 00:31:56 -->
<!-- recording: Sven and Tine 25.01.30 p1; time: 00:00:00 - 00:23:25 -->

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
For instance, a motivating example that illustrates these concepts throughout the document could involve researchers aiming to answer the question: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students to two groups, each receiving a different teaching method. Judges would then compare pairs of students' written texts, producing a dichotomous outcome that reflects the relative quality of each text. Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

<!-- 4. But why identification is so important? -->
However, @sec-theory-issues presents compelling evidence that the BTL model has several statistical and measurement limitations. These limitations hinder the model's ability to identify distinct estimands relevant to CJ inquiries, including the one from the motivating example. Identification is crucial because it is a necessary condition for ensuring consistent estimators. Without identification, researchers cannot achieve *consistency* even with "infinite" and error-free data. In other words, the estimates derived from the estimator would fail to converge to the "true" value of the estimand as the data size approaches infinity [@Everitt_et_al_2010]. As a result, deriving meaningful insights from finite data also becomes unattainable [@Schuessler_et_al_2023, pp. 5].

@@

<!-- 5. What the section will do? -->
As a result, this section addresses the identified limitations by extending Thurstone's general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. The extended model clarifies the interactions among all actors and processes involved in CJ experiments. Furthermore, it allows researchers to shift the current comparative data analysis paradigm from accepting the model's assumptions as given to testing these assumptions for specific comparative data. For details on these assumptions, refer to @sec-theory-issues. Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what the study calls the *conceptual-population model*. Meanwhile, @sec-theory-theoretical_SC integrates the assessment design features into what the study terms the *sample-comparison model*.

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->
<!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->

<!-- 1. The SCMs -->
Before integrating the theoretical principles from @sec-thurstone_theory, the study defines SCMs and DAGs. SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

<!-- 2. the preliminary SCM for CJ -->
@fig-cj04 illustrates the preliminary integration of the theoretical principles underlying the motivating example. Specifically, [SCM -@fig-cj04_scm] displays the relationship between the outcome of the CJ experiment, $O_{abhijk}$, to several other variables. Here, the indices $\{a,b\}$ represent the compared texts, $\{h,i\}$ label the students who wrote these texts, $j$ denotes the judge, and $k$ accounts for cases where the judges perform multiple comparisons of the same texts written by the same students, i.e., the comparison number for a *repeated measures design* [@Lawson_2015, pp. 366-376]. The other variables related to the outcome include the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$. Together with the outcome, these variables form the preliminary set of endogenous variables $V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}$. Finally, the preliminary set of structural equations $F = \{f_{O},f_{D},f_{PT}\}$ defines the dependencies among these variables.

@@

<!-- 3. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where the nodes represent random variables. The term *directed* means that the edges extend from one node to another, with arrows indicating the direction of causal influence. The term *acyclic* implies that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs represent observed variables as solid black circles, while they use open circles for unobserved (latent) variables [@Morgan_et_al_2014]. While the *standard representation* of DAGs typically omits exogenous variables for simplicity, the *magnified representation* includes them, offering the advantage of highlighting potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 4. the preliminary DAG for CJ -->
[DAG -@fig-cj04_dag] displays the preliminary DAG associated with [SCM -@fig-cj04_scm]. The graph shows the direction of causal influences among variables, reflecting the assumptions of Thurstone's theory. It also distinguishes between observable endogenous variables, such as the outcome $O_{abhijk}$, and variables considered latent according to the theory and practice of CJ, such as the discriminal difference $(D_{abhijk})$, the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$.

<!-- Notice,  -->
<!-- the perceived discriminal processes $(PT_{ahjk}, PT_{bijk})$, the "true" discriminal processes $(T_{ahjk}, T_{bijk})$, and the judges' biases $(B_{kj})$ -->

<!-- Because it quickly becomes intractable, and without loosing generality, the (latent) "perceived" and "true" discriminal processes for the stimuli can be depicted in a vector for each judge, as in   -->

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$. The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.
